---
title: "IODS_final"
author: "Oyedayo Oyelowo"
date: "17 December 2017"
output: html_document
---

##Loading the data set and packages
```{r, message=FALSE, warning=FALSE}
#clear environment
rm(list=ls())

#read data
fpath<- "C:/Users/oyeda/Desktop/OPEN_DATA_SCIENCE/IODS-final/data/"
alco_data <- read.table(paste0(fpath, "alcohol_student.csv"), sep=",", header=T)
alco_data<- alco_data[,-1]

#necessary packages
#install.packages("dismo")
library(dplyr)
library(ggplot2)
library(corrplot)
library(GGally)
library(tidyr)
library(tidyverse)
library(gbm)
library(dismo)
library(caTools)
library(mgcv)
library(MASS)
library(FactoMineR)
```


My aim is to demonstrate the use of various statistical techniques in getting insight in to
the data. Firstly, I will use the basic desctiptive statitstics to understand
the distribution, correlation and dependencies(cor, barplot, histogram, biplot etc)
I will also be using the regression models(GLM, GAM and GBM) and test
my models using 70:30 data split.
I will also explore LDA and  MCA.


```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```
```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```




```{r, message=FALSE, warning=FALSE}


```




```{r, message=FALSE, warning=FALSE}


```




```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```
```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```

```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```



```{r, message=FALSE, warning=FALSE}


```


```{r, message=FALSE, warning=FALSE}


```